%
% file: dedication.tex
% author: V?ctor Bre?a-Medina
% description: Contains the text for thesis dedication
%

\chapter*{Dedication and acknowledgements}
\begin{SingleSpace}
\initial{T}his thesis is dedicated to the community of scientists and engineers of all kinds which have made this thesis possible. In particular, however, I would like to acknowledge and thank the following people. 

First, Toni Mey, who has provided so much detailed and substantive feedback on my work, promoted my cause, and been a constant source of encouragement throughout my PhD: from correcting my coding errors on the fly in my first group talk, to last minute analyses in this thesis. To Adrian Mulholland for reading and critiquing this thesis and for support and encouragement throughout my PhD. To Eric Lang for his feedback and useful discussions on chapter \ref{chap:aadh} and to my good friends Fred Cascarini and Emily Lythell for holding my hand through the submission process and for constant emotional support throughout my time in Bristol.  

I would also like to thank the staff at the Jean Golding Institute for providing such a friendly, encouraging and stimulating place to work, in particular: Patty Holly, Natalie Thurlby, Kate-Robson Brown, Rachel Prior and Liz Green. And to the Unite House family of 2018, Sandy Payne, Cait Houlbrook, Felipe C\^{o}rte Real De Camargo and the others: it was emotional. 

Completing this thesis would not have been possibly without the support of my friends and colleagues: Attila Jantsek, Bal\'azs Hornung, M\'{e}langi Panagi George Holloway, Lisa May Thomas, Alex Jones, Tom Mitchell, Mark Wonnacott, Silvia Amabilino, Mike O'Connor, Rob Kent, James Hall,  Helen Murphy, Zara Cassid, Elias \& Rose \& family, Gareth \& Louise \& family, Tom Oliver, and to all those in the Muholland and Glowacki groups. Special thanks must also go to CJ the Wondrous and their partner James for being such amazing friends and providing a wonderful home for me while I write. Lastly my family: my uncle David, cousin Christine, my sister Katie, her husband Mike and of course my mother Susan whose constant, unconditional love and support sustained me throughout this process. 

\end{SingleSpace}
\clearpage